Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-24 paper/submission.tex
Attemt to clarify that causality from sentiment -> price is founded on detecting causality from sentiment changes -> price returns, and vice versa

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
    Information transfer between time series is calculated using the asymmetric information-theoretic measure known as transfer entropy. Geweke's autoregressive formulation of Granger causality is used to compute linear transfer entropy, and Schreiber's general, non-parametric, information-theoretic formulation is used to quantify non-linear transfer entropy. 
    We first validate these measures against synthetic data. Then we apply these measures to detect {\color{blue} statistical causality} between social sentiment and cryptocurrency prices. We validate results by performing permutation tests by shuffling the time series, and calculate the Z-score. We also investigate different approaches for partitioning in nonparametric density estimation which can improve the significance. 
    Using these techniques on sentiment and price data over a 48-month period to August 2018, for four major cryptocurrencies, namely bitcoin (BTC), ripple (XRP), litecoin (LTC) and ethereum (ETH), we detect significant information transfer, on hourly timescales, with greater net information transfer from sentiment to price for XRP and LTC, and instead from price to sentiment for BTC and ETH. We report the scale of non-linear {\color{blue}statistical causality} to be an order of magnitude larger than {\color{blue}the linear case}.
\end{abstract}

@@

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. In the ideal case,  knowing the movement of one price, the movement of a second can be inferred.  In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency prices.

@@

  The relationship between social media sentiment and price has been explored in the literature for traditional markets and, more recently, for crypto markets as well, including by an author of this paper. For instance, Bollen et al.  \cite{bollen2011twitter} showed that the mood of Twitter messages can be used as a proxy for market sentiment, and that this can show a linear relationship with price movements in US equities. Zheludev \& Aste \cite{zheludev2014can} also performed sentiment analysis using Natural Language Processing (NLP) on Twitter data, to show sentiment is significantly coupled with price movements for a number of instruments issued by S\&P500 firms. Souza \& Aste  used Twitter messages to model market sentiment, and showed the non-linear predictive relationship may be greater than the linear one \cite{souza2016nonlinear}.

@@

  Within a linear causality framework, hypothesising that cryptocurrency price depends on prior values of both price and market sentiment, a Granger causality test can detect the impact of past values of a variable $X_t$ on future values of another variable $Y_t$ \cite{granger1969investigating}. This can be calculated using a vector auto-regressive (VAR) model, which describes the extent to which including past values of $X$ reduces the sum of squared residuals in the regression of $X$ against $Y$, hence estimating the predictive effect, at each lag $k$, of the social sentiment at $t-k$ on the price at time $t$.

@@

%\clearpage
\section{Results with Real Data} \label{s.results}
  Having confirmed that the information-theoretic approach is able to detect both linear and non-linear causalities, we apply the technique to investigate the effect of social media sentiment on cryptocurrency prices. We also apply the linear method, and compare estimates from both techniques to establish whether linear or non-linear dynamics dominate the observed causal relationship.

@@

  The choice of timescale in aggregating raw sentiment data involves a trade-off: with too fine a timescale, there are not enough messages to estimate sentiment, but a long timescale represents a low-resolution sample which loses too much information about the underlying time series. Since cryptocurrencies are traded in real time on electronic exchanges, we hypothesise that causal signals between sentiment and price operate at sub-hourly timescales; hourly aggregation is the smallest time period available in the data, and so this aggregation of sentiment is used.

@@

  For BTC, in Fig. \ref{fig:BTC_TE}, we detect a strong causative signal, of roughly similar scale in both directions of sentiment to BTC price and in the reverse direction. Analysis of the Z-scores confirms the significance in both directions, with the net information transfer greater in the direction of price to sentiment, with mean net transfer entropy of $-0.0016$ bits.

  LTC, in Fig. \ref{fig:LTC_TE}, shows a similar pattern to BTC, although in this case the predominant direction of information transfer is reversed, with mean net transfer entropy of $0.0039$ bits. The Z-scores show the significance of sentiment to price is consistently greater than in the reverse direction, and also stronger than the other currencies. In addition, there is notably a large linear component to the information transfer, which is unique to LTC.

  XRP, in Fig. \ref{fig:XRP_TE}, shows a clear non-linear causality from sentiment to price, with a significant but smaller causal relationship also in the opposite direction. The mean net transfer entropy is measured for this period as $0.0022$ bits. The signal is more significant from sentiment to price, and especially in the periods ending in 2018.

  ETH, in Fig. \ref{fig:ETH_TE}, shows an interesting and unique behaviour. In particular, there appears to be, initially, a significant signal which collapses in the windows ending around January 2018. The information transfer at its initial peak is greater from price to sentiment, however over the whole period the effect is more bidirectional, with mean net transfer entropy of zero. This suggests a slight phase change, with sentiment driving price after the collapse of the signal, although no net significance is observed after this point in either direction. This strongly indicates another driving mechanism, the effect of which first becomes present around January 2016 (due to 24-month windows). This effect is likely to be associated with the rapid price movements at the time.  %\cite{Ethereum}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/BTC.pdf}
    \caption{Evidence that BTC sentiment and price are causally coupled in both directions in a non-linear way. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a high level of significance, especially during 2017 and 2018, in both directions, with the peak significance $Z=21.3$ observed in the direction of price to sentiment in the 24-month window ending February 2018.}
    \label{fig:BTC_TE}

    \includegraphics[width=\linewidth]{images/LTC.pdf} 
    \caption{Evidence that LTC price and sentiment are causally coupled in both directions in a non-linear way, with sentiment having a larger influence on price than the other way round. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a significant signal in both directions, with the net information transfer generally operating in the direction of sentiment to price. Peak significance of $Z=42.8$ is observed from sentiment to price in the 24-month window ending February 2018.}
    \label{fig:LTC_TE}
    \vspace{-4pt}
  \end{figure*}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/XRP.pdf}
    \caption{Evidence that XRP price and sentiment are causally coupled in both directions in a non-linear way, with the prevailing direction of information transfer flowing from sentiment to price in the first period, and from price to sentiment in the second. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a small but clear significant signal, in both directions, which decays rapidly towards January 2018 and does not recover afterward. Peak significance of $Z=41.9$ is observed from sentiment to price in the 24-month window ending January 2018.}
    \label{fig:XRP_TE}

    \includegraphics[width=\linewidth]{images/ETH.pdf} 
    \caption{Evidence that ETH price and sentiment are causally coupled in both directions in a non-linear way. Overall this coupling is of lower significance compared to the other currencies investigated. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, indicate initial significance for windows ending before 2018, followed by low significance after the collapse in signal strength, which must therefore begin around January 2016. Peak significance is observed in the first window of the study, with $Z=20.6$ in the direction of sentiment to price, and $Z=19.9$ in the direction of price to sentiment.     
  }
  \label{fig:ETH_TE}
  \vspace{-34pt}
  \end{figure*}

@@

Information-theoretic and autoregressive techniques were developed and validated on coupled random walks and chaotic logistic maps, confirming the ability of both techniques to detect linear information transfer, and of the information-theoretic technique to detect non-linear information transfer. Following validation, the techniques were applied to historical data describing social media sentiment and cryptocurrency prices to detect information transfer between sentiment and price movements.

The information-theoretic investigation detected a significant non-linear causal relationship in BTC, LTC and XRP, over multiple timescales and in both directions of sentiment to price and price to sentiment. The effect was strongest and most consistent for LTC and XRP, and in both cases the net information transfer was in the direction of sentiment to price. ETH and BTC were observed to show significant net information transfer in the direction price to sentiment, although separate dynamics were observed in ETH which see a collpase in significance. {\color{blue}<removed>}%Given the hypothesis that low barriers to entry and unsophisticated investor speculation are key drivers for price movements, and that BTC and ETH represent the most widely known and traded cryptocurrencies, the fact that causality is detected most clearly for these currencies corresponds to expectations.

@@

  \begin{table*}[!htb]
    \label{t.table3}
    \caption{ Net information transfer statistics in units of bits determined over all windows of the study for each cryptocurrency. Mean values show that in time, more information was transfered from sentiment to price for LTC and XRP, with more information transfer in the direction of price to sentiment for BTC and ETH. It should be noted that mean due to the overlapping strides of the windows, periods during the middle of the sample are weighted more heavily; however the figures appropriately present the relative direction of information transfer over the 48-month period of the study, so are presented for the sake of comparison. Significance scores are presented in Table {\color{blue}4}.}

@@

    \begin{tabular}{lrrrr}
      {} & {} & (Price $\rightarrow$ Sentiment) & {} & (Sentiment $\rightarrow$ Price)  \\
      {} & Linear  &  NonLinear  &  Linear  &  Nonlinear  \\
      \hline 
      % BTC &     0.1 &     2.0 &        5.1 &        7.6 \\
      % LTC &     2.2 &    11.8 &       22.4 &       17.3 \\
      % XRP &     4.1 &     0.4 &       24.1 &       14.4 \\
      % ETH &    -0.6 &     0.5 &        6.5 &        6.8 \\
      BTC &     0.1 &       5.1 &       2.0 &        7.6 \\
      LTC &     2.2 &      22.4 &      11.8 &       17.3 \\
      XRP &     4.1 &      24.1 &       0.4 &       14.4 \\
      ETH &    -0.6 &       6.5 &       0.5 &        6.8 \\
      \end{tabular}

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
    Information transfer between time series is calculated using the asymmetric information-theoretic measure known as transfer entropy. Geweke's autoregressive formulation of Granger causality is used to compute linear transfer entropy, and Schreiber's general, non-parametric, information-theoretic formulation is used to quantify non-linear transfer entropy. 
    We first validate these measures against synthetic data. Then we apply these measures to detect {\color{blue} statistical causality} between  social sentiment {\color{blue} changes} and cryptocurrency {\color{blue} returns}. We validate results by performing permutation tests by shuffling the time series, and calculate the Z-score. We also investigate different approaches for partitioning in nonparametric density estimation which can improve the significance. 
    Using these techniques on sentiment and price data over a 48-month period to August 2018, for four major cryptocurrencies, namely bitcoin (BTC), ripple (XRP), litecoin (LTC) and ethereum (ETH), we detect significant information transfer, on hourly timescales, with greater net information transfer from sentiment to price for XRP and LTC, and instead from price to sentiment for BTC and ETH. We report the scale of non-linear {\color{blue}statistical causality} to be an order of magnitude larger than {\color{blue}the linear case}.
\end{abstract}

@@

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. In the ideal case, {\color{blue}from} knowing the movement of one price, the movement of a second can be inferred.  In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency {\color{blue} price returns}.

@@

  The relationship between social media sentiment and price {\color{blue} changes} has been explored in the literature for traditional markets and, more recently, for crypto markets as well, including by an author of this paper. For instance, Bollen et al.  \cite{bollen2011twitter} showed that the mood of Twitter messages can be used as a proxy for market sentiment, and that this can show a linear relationship with price movements in US equities. Zheludev \& Aste \cite{zheludev2014can} also performed sentiment analysis using Natural Language Processing (NLP) on Twitter data, to show sentiment is significantly coupled with price movements for a number of instruments issued by S\&P500 firms. Souza \& Aste  used Twitter messages to model market sentiment, and showed the non-linear predictive relationship may be greater than the linear one \cite{souza2016nonlinear}.

@@

  Within a linear causality framework, hypothesising that cryptocurrency price {\color{blue} returns depend} on prior values of both {\color{blue} price returns} and market sentiment, a Granger causality test can detect the impact of past values of a variable $X_t$ on future values of another variable $Y_t$ \cite{granger1969investigating}. This can be calculated using a vector auto-regressive (VAR) model, which describes the extent to which including past values of $X$ reduces the sum of squared residuals in the regression of $X$ against $Y$, hence estimating the predictive effect, at each lag $k$, of the {\color{blue} change in} social sentiment at $t-k$ on the {\color{blue} returns} at time $t$.

@@

%\clearpage
\section{Results with Real Data} \label{s.results}
  Having confirmed that the information-theoretic approach is able to detect both linear and non-linear causalities, we apply the technique to investigate the effect of {\color{blue} changes in} social media sentiment on cryptocurrency {\color{blue} returns, and vice versa}. We also apply the linear method, and compare estimates from both techniques to establish whether linear or non-linear dynamics dominate the observed causal relationship.

@@

  The choice of timescale in aggregating raw sentiment data involves a trade-off: with too fine a timescale, there are not enough messages to estimate sentiment, but a long timescale represents a low-resolution sample which loses too much information about the underlying time series. Since cryptocurrencies are traded in real time on electronic exchanges, we hypothesise that causal signals between sentiment and {\color{blue} returns} operate at sub-hourly timescales; hourly aggregation is the smallest time period available in the data, and so this aggregation of sentiment is used.

@@

  For BTC, in Fig. \ref{fig:BTC_TE}, we detect a strong causative signal, of roughly similar scale in both directions of sentiment to BTC {\color{blue} returns} and in the reverse direction. Analysis of the Z-scores confirms the significance in both directions, with the net information transfer greater in the direction of {\color{blue} returns} to sentiment, with mean net transfer entropy of $-0.0016$ bits.

  LTC, in Fig. \ref{fig:LTC_TE}, shows a similar pattern to BTC, although in this case the predominant direction of information transfer is reversed, with mean net transfer entropy of $0.0039$ bits. The Z-scores show the significance of sentiment to {\color{blue} returns} is consistently greater than in the reverse direction, and also stronger than the other currencies. In addition, there is notably a large linear component to the information transfer, which is unique to LTC.

  XRP, in Fig. \ref{fig:XRP_TE}, shows a clear non-linear causality from sentiment to {\color{blue} returns}, with a significant but smaller causal relationship also in the opposite direction. The mean net transfer entropy is measured for this period as $0.0022$ bits. The signal is more significant from sentiment to {\color{blue} returns}, and especially in the periods ending in 2018.

  ETH, in Fig. \ref{fig:ETH_TE}, shows an interesting and unique behaviour. In particular, there appears to be, initially, a significant signal which collapses in the windows ending around January 2018. The information transfer at its initial peak is greater from {\color{blue} returns} to sentiment, however over the whole period the effect is more bidirectional, with mean net transfer entropy of zero. This suggests a slight phase change, with sentiment driving {\color{blue} returns} after the collapse of the signal, although no net significance is observed after this point in either direction. This strongly indicates another driving mechanism, the effect of which first becomes present around January 2016 (due to 24-month windows). This effect is likely to be associated with the rapid price movements at the time.  %\cite{Ethereum}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/BTC.pdf}
    \caption{Evidence that BTC sentiment and {\color{blue} returns} are causally coupled in both directions in a non-linear way. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a high level of significance, especially during 2017 and 2018, in both directions, with the peak significance $Z=21.3$ observed in the direction of {\color{blue} returns} to sentiment in the 24-month window ending February 2018.}
    \label{fig:BTC_TE}

    \includegraphics[width=\linewidth]{images/LTC.pdf} 
    \caption{Evidence that LTC {\color{blue} returns} and sentiment are causally coupled in both directions in a non-linear way, with sentiment having a larger influence on {\color{blue} returns} than the other way round. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a significant signal in both directions, with the net information transfer generally operating in the direction of sentiment to {\color{blue} returns}. Peak significance of $Z=42.8$ is observed from sentiment to {\color{blue} returns} in the 24-month window ending February 2018.}
    \label{fig:LTC_TE}
    \vspace{-4pt}
  \end{figure*}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/XRP.pdf}
    \caption{Evidence that XRP {\color{blue} returns} and sentiment are causally coupled in both directions in a non-linear way, with the prevailing direction of information transfer flowing from sentiment to {\color{blue} returns} in the first period, and from {\color{blue} returns} to sentiment in the second. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a small but clear significant signal, in both directions, which decays rapidly towards January 2018 and does not recover afterward. Peak significance of $Z=41.9$ is observed from sentiment to {\color{blue} returns} in the 24-month window ending January 2018.}
    \label{fig:XRP_TE}

    \includegraphics[width=\linewidth]{images/ETH.pdf} 
    \caption{Evidence that ETH {\color{blue} returns} and sentiment are causally coupled in both directions in a non-linear way. Overall this coupling is of lower significance compared to the other currencies investigated. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, indicate initial significance for windows ending before 2018, followed by low significance after the collapse in signal strength, which must therefore begin around January 2016. Peak significance is observed in the first window of the study, with $Z=20.6$ in the direction of sentiment to {\color{blue} returns}, and $Z=19.9$ in the direction of {\color{blue} returns} to sentiment.     
  }
  \label{fig:ETH_TE}
  \vspace{-34pt}
  \end{figure*}

@@

Information-theoretic and autoregressive techniques were developed and validated on coupled random walks and chaotic logistic maps, confirming the ability of both techniques to detect linear information transfer, and of the information-theoretic technique to detect non-linear information transfer. Following validation, the techniques were applied to historical data describing social media sentiment and cryptocurrency prices to detect information transfer between {\color{blue} changes in} sentiment and price {\color{blue} returns}.

The information-theoretic investigation detected a significant non-linear causal relationship in BTC, LTC and XRP, over multiple timescales and in both directions of sentiment to {\color{blue} returns} and {\color{blue} returns} to sentiment. The effect was strongest and most consistent for LTC and XRP, and in both cases the net information transfer was in the direction of sentiment to {\color{blue} returns}. ETH and BTC were observed to show significant net information transfer in the direction {\color{blue} returns} to sentiment, although separate dynamics were observed in ETH which see a collapse in significance. {\color{blue}<sentence removed>}%Given the hypothesis that low barriers to entry and unsophisticated investor speculation are key drivers for price movements, and that BTC and ETH represent the most widely known and traded cryptocurrencies, the fact that causality is detected most clearly for these currencies corresponds to expectations.

@@

  \begin{table*}[!htb]
    \label{t.table3}
    \caption{ Net information transfer statistics in units of bits determined over all windows of the study for each cryptocurrency. Mean values show that in time, more information was transfered from sentiment to price for LTC and XRP, with more information transfer in the direction of price to sentiment for BTC and ETH. It should be noted that due to the overlapping strides of the windows, periods during the middle of the sample are weighted more heavily; however the figures appropriately present the relative direction of information transfer over the 48-month period of the study, so are presented for the sake of comparison. Significance scores are presented in Table {\color{blue}4}.}

@@

    \begin{tabular}{lrrrr}
      {} & {} & ({\color{blue} returns} $\rightarrow$ Sentiment) & {} & (Sentiment $\rightarrow$ {\color{blue} returns})  \\
      {} & Linear  &  NonLinear  &  Linear  &  Nonlinear  \\
      \hline 
      % BTC &     0.1 &     2.0 &        5.1 &        7.6 \\
      % LTC &     2.2 &    11.8 &       22.4 &       17.3 \\
      % XRP &     4.1 &     0.4 &       24.1 &       14.4 \\
      % ETH &    -0.6 &     0.5 &        6.5 &        6.8 \\
      BTC &     0.1 &       5.1 &       2.0 &        7.6 \\
      LTC &     2.2 &      22.4 &      11.8 &       17.3 \\
      XRP &     4.1 &      24.1 &       0.4 &       14.4 \\
      ETH &    -0.6 &       6.5 &       0.5 &        6.8 \\
      \end{tabular}
